Typed Clojure: A Practical Optional Type System for Clojure

### 2012-10-02 dissertation/design-choices.tex
Experiments

Typed Clojure is designed to be of practical use to Clojure programmers.
Many of the design goals are similar to Tobin-Hochstadt's \cite{SAMTH:dissertation}
for Typed Racket.

@@

\section{Optional Type Checking}
\section{Local Type Inference}

@@

$$
\begin{tdisplay}{Evaluation Contexts}
  \begin{altgrammar}
    \E{} &::=& % application rules
              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
              % map rules
              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
              &\mbox{Evaluation Contexts}
  \end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Operational Semantics}
\begin{array}{c}
{\inferrule[E-Assoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
    {{(\assoc{}\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
\\
{\inferrule[E-Dissoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
  {(dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
\\
{\inferrule[E-GetMapExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_1\ has\ entry\ with\ key\ v_2} \\
    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1} }
  { {(get\ v_1\ v_2) \hookrightarrow v_3} }} \\
\\
{\inferrule[E-GetMapNotExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_1\ has\ no\ entry\ with\ key\ v_2} \\
    {v_3 = \nil{}} }
  { {(get\ v_1\ v_2) \hookrightarrow v_3} }}
\end{array} 
\end{tdisplay} 
$$

$$
\begin{tdisplay}{Type Syntax}
\begin{altgrammar}
  \T{} &::=& \Top \alt \Bottom \alt \Nil \alt \k{} \alt \{ \overrightarrow{\k{}\ \T{}} \}
             \alt {\IPersistentMap {\T{}} {\T{}}} \alt ( \cup\ \overrightarrow{\T{}} )
             \alt ( \cap\ \overrightarrow{\T{}} )
\end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Core Type Rules}
\begin{array}{c}
{\inferrule[T-AssocHMap]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\k{}}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
    {\vdash {\issubtype {{\T{}}_{m}} {{ \{ \} }} }} \\
    {\T{} = (update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})} }
  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
\\
{\inferrule[T-AssocHMapPromote]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
    {\vdash {\issubtype {{\T{}}_{m}} { \{ \}}}} \\
    {\T{} = (promote\_hmap\ {\T{}}_{m}\ {\T{}}_{k}\ {\T{}}_{v})} }
  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
\\
{\inferrule[T-AssocIPersistentMap]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}} }} \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}} }} \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}} }} }
  { {\Gamma \vdash {\hastype {(\assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {T}}}}}
\end{array}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Type Metafunctions}
\begingroup
\fontsize{10pt}{12pt}
\begin{altgrammar}
  (update\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\k{}}_1\ {\T{}}_1) = 
                 \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}}\ {\k{}}_1\ {\T{}}_1 \}\ \\
  (update\_hmap\ (\cap\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v} = 
                               (\cap\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})} \\
  (update\_hmap\ (\cup\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v} = 
                               (\cup\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
  \\ \\
  (promote\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\T{}}_{kn}\ {\T{}}_{vn})
                                    = {\IPersistentMap {(\cup\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
                                                       {(\cup\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
  (promote\_hmap\ {\IPersistentMap {{\T{}}_{k}} {{\T{}}_{v}}} \ {\T{}}_{kn}\ {\T{}}_{vn})
                                    = {\IPersistentMap {(\cup\ {\T{}}_{k}\ {\T{}}_{kn})}
                                                       {(\cup\ {\T{}}_{v}\ {\T{}}_{vn})}}
\end{altgrammar}
\endgroup
\end{tdisplay}
$$

\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}

Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
representing a sequence of types, allowing non-uniform variable parameters.
This enables \lstinline|map| and other function with complex arguments to be typed.
Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
Typed Racket, and is a key feature of Typed Racket's implementation.

Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
associates key-value pairs to a target map. It takes one or more key-value pairs and
applies them left-to-right to the target map. The key problem from a typing perspective
is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
higher-order function.
\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
dotted parameters represent an unrestricted number of types, where we require an even number.

The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
Firstly, subtyping between different kinds of dotted and rest
\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
difficult to express.
Thirdly, higher-order functions taking functions with dotted arguments commonly 
The first issue requires careful design and implementation.
The second and third points require new constructs, in the form of \emph{projections}
and abstractions over \emph{dotted} parameters, respectively.

>>>>

Typed Clojure is designed to be of practical use to Clojure programmers.
Many of the design goals are similar to Tobin-Hochstadt's \cite{Tob10}
for Typed Racket.

@@

\section{Multimethods}

% INTRODUCTION TO CLOJURE MULTIMETHODS

%A multimethod in Clojure differs from most other
%abstractions of the same name. Clojure multimethods
%dispatch on the result of an arbitrary function
%rather than the types or structure of its arguments.
%
%\begin{lstlisting}[caption=Clojure Multimethods, label=lst:mm1]
%(defmulti ToString class)
%
%(defmethod ToString java.lang.Integer
%  [a]
%  (str "Integer: " a))
%
%(defmethod ToString java.lang.Number
%  [a]
%  (str "Number: " a))
%\end{lstlisting}
%
%Even with this extra flexibility, multimethods are often used to dispatch on the type of its arguments.
%Listing \ref{lst:mm1} illustrates a common usage of multimethods: dispatching on the class of the first argument.
%We use \lstinline|defmulti| to define a new multimethod.
%The function \lstinline|class| is used as the dispatch function, a one-argument function returning the class of its argument
%as an instance of \lstinline|java.lang.Class|.
%
%A multimethod invocation redirects control flow to exactly one of its \emph{methods} with unresolved ambiguities resulting in
%a runtime error. The method is selected by first comparing the result of the dispatch function to
%the dispatch values defined by each method. If one or more methods match, multimethods provide
%the ability to prefer one method over another.

\section{Clojure Class Hierarchy}
\section{Protocols and Datatypes}
\section{Optional Type Checking}
\section{Local Type Inference}

@@

$$
\begin{tdisplay}{Evaluation Contexts}
  \begin{altgrammar}
    \E{} &::=& [ ] % application rules
              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
              % map rules
              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
              &\mbox{Evaluation Contexts}
  \end{altgrammar}
\end{tdisplay}
$$

% \mapsto for updating maps
$$
\begin{tdisplay}{Operational Semantics}
\begin{array}{c}
{\inferrule[E-Assoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
    {{(\assoc{}\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
\\
{\inferrule[E-Dissoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
  {(dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
\\
{\inferrule[E-GetMapExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_2\ v_3\ in\ v_1} }
%    {v_1\ has\ entry\ with\ key\ v_2} \\
%    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1}
  { {(get\ v_1\ v_2) \hookrightarrow v_3} }} \\
\\
{\inferrule[E-GetMapNotExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_1\ has\ no\ entry\ with\ key\ v_2} \\
    {v_3 = \nil{}} }
  { {(get\ v_1\ v_2) \hookrightarrow v_3} }}
\end{array} 
\end{tdisplay} 
$$

$$
\begin{tdisplay}{Type Syntax}
\begin{altgrammar}
  \T{} &::=& \Top \alt \Bottom \alt \Nil \alt \k{} \alt \{ \overrightarrow{\k{}\ \T{}} \}
             \alt {\IPersistentMap {\T{}} {\T{}}} \alt ( \vee\ \overrightarrow{\T{}} )
             \alt ( \wedge\ \overrightarrow{\T{}} )
\end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Core Type Rules}
\begin{array}{c}
{\inferrule[T-AssocHMap]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\k{}}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
    {\vdash {\issubtype {{\T{}}_{m}} {\TopHMap} }} }
  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} 
        {(update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})}}}}} \\
\\
{\inferrule[T-AssocPromote]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
    {\vdash {\issubtype {{\T{}}_{m}} {\TopIPersistentMap}}} \\
    {\T{} = (promote\_hmap\ {\T{}}_{m}\ {\T{}}_{k}\ {\T{}}_{v})} }
  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
\end{array}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Type Metafunctions}
\begingroup
\fontsize{10pt}{12pt}
\begin{altgrammar}
  (update\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\k{}}_1\ {\T{}}_1) = 
                 \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}}\ {\k{}}_1\ {\T{}}_1 \}\ \\
  (update\_hmap\ (\wedge\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
                               (\wedge\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})}) \\
  (update\_hmap\ (\vee\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
                               (\vee\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
  \\ \\
  (promote\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\T{}}_{kn}\ {\T{}}_{vn})
                                    = {\IPersistentMap {(\vee\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
                                                       {(\vee\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
  (promote\_hmap\ {\IPersistentMap {{\T{}}_{k}} {{\T{}}_{v}}} \ {\T{}}_{kn}\ {\T{}}_{vn})
                                    = {\IPersistentMap {(\vee\ {\T{}}_{k}\ {\T{}}_{kn})}
                                                       {(\vee\ {\T{}}_{v}\ {\T{}}_{vn})}}
\end{altgrammar}
\endgroup
\end{tdisplay}
$$

$$
\begin{tdisplay}{Subtyping}
\begin{array}{c}
\inferrule[S-Refl] { } {\vdash {\issubtype {\T{}} {\T{}}}} \\
\inferrule[S-Any] { } {\vdash {\issubtype {\T{}} {\Top}}}
\inferrule[S-UnionSuper] 
  {\ } 
  {\vdash {\issubtype {\T{}} {\Top}}}
\end{array}
\end{tdisplay}
$$

%\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}
%
%Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
%representing a sequence of types, allowing non-uniform variable parameters.
%This enables \lstinline|map| and other function with complex arguments to be typed.
%Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
%Typed Racket, and is a key feature of Typed Racket's implementation.
%
%Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
%to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
%variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
%associates key-value pairs to a target map. It takes one or more key-value pairs and
%applies them left-to-right to the target map. The key problem from a typing perspective
%is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
%accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
%higher-order function.
%\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
%dotted parameters represent an unrestricted number of types, where we require an even number.
%
%The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
%new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
%Firstly, subtyping between different kinds of dotted and rest
%\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
%parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
%Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
%difficult to express.
%Thirdly, higher-order functions taking functions with dotted arguments commonly 
%The first issue requires careful design and implementation.
%The second and third points require new constructs, in the form of \emph{projections}
%and abstractions over \emph{dotted} parameters, respectively.
